A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-10 paper.tex
Move arg inference methods to appendix

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies, leading to basic errors in statements
about what different inference methods produce as output (see the
``\nameref{ARG_inference_methods}'' section), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

\section*{ARG inference methods}\label{ARG_inference_methods}
Using the terminology developed here to classify ARGs, we now review methods developed
to explicitly \emph{infer} ARGs from sequencing data.
% Gregor: I expected clear indication which method/tool works with eARGs and gARGs, but I can see this clearly

\subsection*{ARG reconstruction}

% AI: just to note, these papers do not all agree in their definition of an ARG.
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?

The likelihood $p(D | G)$ is effectively always the distribution of
a Poisson process on the ancestral material along the edges of $G$ \citep[Eq.\ (2)]{mahmoudi2022bayesian}.
Hence it is typically straightforward to evaluate in principle, but choosing the
right data structure can still have a dramatic effect on the efficiency of evaluation
\citep{mahmoudi2022bayesian}. Especially in the Bayesian case, $p(G)$ is typically specified
as the sampling distribution of a stochastic process, such as the coalescent with
recombination. Two contemporary examples are
\cite{mahmoudi2022bayesian, guo2022recombination}.
Evaluating the distribution of the coalescent with recombination
(or related models) requires knowledge of the number of extant lineages and
number of links available for recombination (i.e.\ ones at which a recombination
would split ancestral material) at each time \citep[Eq.\ (3)]{mahmoudi2022bayesian}.
These require a data structure from which shared edges on different trees are easy to identify,
and which encodes recombination events and times explicitly, e.g.\ using nodes.
We will provide a more detailed discussion on sampling probability evaluation in the appendix.

@@

\end{document}

>>>>

Recent breakthroughs in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,wohns2022unified}
and data representation~\citep{kelleher2016efficient}
have raised the realistic prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using these inferred ancestries as input have
begun to
appear~\citep{osmond2021estimating,zhang2021biobank,fan2022genealogical}
and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies, leading to basic errors in statements
about what different inference methods produce as output, as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a user's perspective) using multiple inference methods in an analysis.

@@

@@

\subsection*{Survey of ARG inference methods}

% AI: just to note, these papers do not all agree in their definition of an
% ARG. I guess this is one of the points of the present paper, but should we
% mention this fact? Should we also mention that most of these infer just the
% topology, but some also the times?

The likelihood $p(D | G)$ is effectively always the distribution of a Poisson
process on the ancestral material along the edges of $G$ \citep[Eq.\
(2)]{mahmoudi2022bayesian}. Hence it is typically straightforward to evaluate
in principle, but choosing the right data structure can still have a dramatic
effect on the efficiency of evaluation \citep{mahmoudi2022bayesian}. Especially
in the Bayesian case, $p(G)$ is typically specified as the sampling
distribution of a stochastic process, such as the coalescent with
recombination. Two contemporary examples are \cite{mahmoudi2022bayesian,
guo2022recombination}. Evaluating the distribution of the coalescent with
recombination (or related models) requires knowledge of the number of extant
lineages and number of links available for recombination (i.e.\ ones at which a
recombination would split ancestral material) at each time \citep[Eq.\
(3)]{mahmoudi2022bayesian}. These require a data structure from which shared
edges on different trees are easy to identify, and which encodes recombination
events and times explicitly, e.g.\ using nodes.
We will provide a more detailed discussion on sampling probability evaluation in the appendix.
\end{document}
